External priors for the next generation of CMB experiments.

### 2015-10-12 Notes/N_eff_prior_paper.tex
new paper text label v2

\date{\today}
\begin{abstract}
The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. 
Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 experiment.
We find that CMB S4... 
\end{abstract}

@@

In this paper we want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
We start this section in \refssec{fm-formalism} by introducing the Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints. To apply this technique, we need to specify our cosmological model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the specifications of S4 CMB experiments. These will be presented in \refssec{cosmo-noise}.

To estimate errors on cosmological parameters we follow the Bayes theorem that relates the likelihood to measure a set of data given the parameters of the model $\mathcal{L}(d|\theta)$, to what we want: the posterior probability of those parameters given the data, $\mathcal{L}(\theta|d)$.
These two are related by the prior probability of the parameters $P(\theta)$ through:
\begin{equation}
\mathcal{L}(\theta|d)\propto \mathcal{L}(d|\theta)P(\theta),
\end{equation}
where, as usual, we have neglected the probability of the data $P(d)$.
These general framework can be simplified in our case. We will focus not on an entirely general posterior distribution, but we will assume that the likelihoods are gaussian.
Furthermore we will not derive the global shape of the likelihood but we will obtain parameters errors by studying small perturbations around its maximum. These are the two basic assumptions of the Fisher-matrix approach.

Regarding the first assumption, we recognize the fact that, even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}. Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.

Secondly we will assume to know the true ``fiducial'' parameters that maximize the likelihood $\mathcal{L}(\theta|d)$ and we will get the errors on those from the likelihood curvature around the fiducial values.
Indeed, as usual, we define the fisher matrix elements as the curvature:
\begin{equation}
	\centering
		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle,
	\label{eqn:Fij_def}
\end{equation}
where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximizes the likelihood.
 With the usual definition \cite{} $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
\begin{equation}
 F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right).
 \label{eqn:Fij_def2}
 \end{equation}
 In this work we will constrain cosmological parameters with the CMB temperature and E mode polarization together with the reconstructed lensing potential. For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
 \begin{eqnarray}
 	\centering
		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
	\label{eqn:cov_definition}
\end{eqnarray}
Note that we are neglecting the term $C_\ell^{E\phi}$. As also noticed in previous literature (like \cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
The term $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
The power of the Fisher approach descend from the Cramer-Rao inequality that relates the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
\begin{equation}
\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}.
\label{eqn:cramer-rao}
\end{equation}
Once $F_{ij}$ is computed following \refeq{Fij_def2} it is straightforwards to get the error $\sigma_i$ from \refeq{cramer-rao}. Furthermore in this context it is easy to introduce external priors on cosmological parameters.
Indeed to introduce priors we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
\begin{equation}
F_{\rm total}=F_{\rm CMB}+\sum F_{\rm external}.
\end{equation}
In the same way we can add a prior on a single cosmological parameter just by modyfing the correspondent matrix element.
For example, a $1\%$ prior on $H_{0}$ can be obtained by:
\begin{equation}
F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,\text{fid}})^2}.
\end{equation}

@@

In the standard cosmology, three active neutrinos are thermally produced in the early universe. Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation. However decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$. The additional fraction depends not only on well-known neutrino scattering rates but also finite temperature quantum corrections. 
Upcoming experiment have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
Furthermore any possible significant deviation from this value could be a hint of a different scenario not predicted by the standard model. 
%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_massless_neutrinos_snow_mass.pdf}
\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the other 6 cosmological parameters of the standard $\Lambda$CDM model. Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{\rm old}=4.1\%$. The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$. The bottom-most (grey) curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
\label{fig:prior_massless_neutrinos}
\end{center}
\end{figure}
%==========

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_omnuh2_snow_mass.pdf}
\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.  Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$. The lower limit on the sum of the neutrino masses is 50 meV.}
\label{fig:prior_omeganuh2}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_w_snow_mass.pdf}
\caption{Constraints on the dark energy equation of state $w$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}. An external prior on $H_{0}$ will be crucial to improve the constraint.} 
\label{fig:prior_w}
\end{center}
\end{figure}

>>>>

\date{\today}
\begin{abstract}
The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. 
Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$  
if the largest scale at reach is not enough to constraint $\tau$ internally. An external prior on the Hubble constant can improve the CMB S4 constraint significantly while complementary information on $n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
\end{abstract}

@@

We want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
We approach the problem with a Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints. To apply this technique, we need to specify our cosmological model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the specifications of S4 CMB experiments.

As usual, we define the fisher matrix elements as the curvature:
\begin{equation}
	\centering
		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle,
	\label{eqn:Fij_def}
\end{equation}
where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximizes the likelihood. It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian. Even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}. Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.

With the usual definition $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
\begin{equation}
 F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right).
 \label{eqn:Fij_def2}
 \end{equation}
 In this work we will constrain cosmological parameters with the CMB temperature and E mode polarization together with the reconstructed lensing potential. For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
 \begin{eqnarray}
 	\centering
		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
	\label{eqn:cov_definition}
\end{eqnarray}
Note that we are neglecting the term $C_\ell^{E\phi}$. As also noticed in previous literature (like \cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
The term $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
The power of the Fisher approach descend from the Cramer-Rao inequality that relates the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
\begin{equation}
\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}.
\label{eqn:cramer-rao}
\end{equation}
Once $F_{ij}$ is computed following \refeq{Fij_def2} it is straightforwards to get the error $\sigma_i$ from \refeq{cramer-rao}. Furthermore in this context it is easy to introduce external priors on cosmological parameters.
Indeed to introduce priors we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
\begin{equation}
F_{\rm total}=F_{\rm CMB}+\sum F_{\rm external}.
\end{equation}
In the same way we can add a prior on a single cosmological parameter just by modyfing the correspondent matrix element.
For example, a $1\%$ prior on $H_{0}$ can be obtained by:
\begin{equation}
F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,\text{fid}})^2}.
\end{equation}

@@

In the standard cosmology, three active neutrinos are thermally produced in the early universe. Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation. However decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$. The additional fraction depends not only on well-known neutrino scattering rates but also finite temperature quantum corrections. 
Upcoming experiment have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be an amazing test of our understanding of the Universe when it was about a second old.
Furthermore any possible significant deviation from this value could be a hint of a different scenario not predicted by the standard model. 
%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_massless_neutrinos_snow_mass_lmin=4_lmax=4499.pdf}
\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of the strength of the prior on the other 6 cosmological parameters of the standard $\Lambda$CDM model. Each prior is expressed in units of the error that would be obtained internally from our suite of CMB experiments; e.g., $\sigma(H_0)_{\rm old}=4.1\%$. The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$. The bottom-most (grey) curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.}
\label{fig:prior_massless_neutrinos}
\end{center}
\end{figure}
%==========

@@

%==========

\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.  Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$. The lower limit on the sum of the neutrino masses is 50 meV.}
\label{fig:prior_omeganuh2}
\end{center}
\end{figure}
%==========

@@

%==========

\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_w_snow_mass_lmin=4_lmax=4499.pdf}
\caption{Constraints on the dark energy equation of state $w$ as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}. An external prior on $H_{0}$ will be crucial to improve the constraint.} 
\label{fig:prior_w}
\end{center}
\end{figure}
%==========
